---
title: "Styleized Streets"
description: |
  A short description of the post.
author:
  - name: mxblsdl
date: 2022-01-30
output:
  distill::distill_article:
    self_contained: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

The following work flow is inspired by some posts I saw on Reddit visualizing the streets of major cities by their suffixes. I decided to do a version for Portland. I came up with my own workflow for creating these stylized maps with R.

Data for streets of Portland can be found [on their open data website](https://gis-pdx.opendata.arcgis.com/datasets/streets). This will differ for other cities but most major cities should have open data for various civic categories.

To find the street data I search on the keyword 'centerline' and found what I was looking for quickly.

## Load Libraries

```{r results='hide', warning=F, message=FALSE}
library(sf) # for spatial manipulation
library(dplyr) # data wrangling
library(ggplot2) # visualization and final graphing
library(leaflet) # interactive widget library
# library(widgetframe) # helps rendering leaflet in markdown
# library(cartography) # color palettes
```

## Find and Load Data

You could download the data and source it locally, but the `sf` package makes it easy to interact with a geojson API. Simply point to the endpoint and load as an sf object.

```{r download-data}
roads <- read_sf("https://opendata.arcgis.com/datasets/ad5ed4193110452aac2d9485df3298e2_68.geojson")
# inspect to find columns of interest
# head(roads)
```

It is almost guaranteed that a spatial file you find from an ArcGIS API will have tons of fields you don't need. I like to drop anything that I won't need for further analysis. 

After inspecting the data, I find the `FTTYPE` column. This step will take some digging with different datasets.

## Prepare Data

Select only the relevant columns for plotting

```{r}
roads <-
  roads %>%
  select(FTYPE)
```

Now we need to create the canvas or boundaries of the final plot. City boundaries can be very odd and hard to capture so I'm just going to use a circle around the city center.

Now we select a point to center the map around. I played around with this a bit using [Google Earth](https://earth.google.com/web/) and chose a point in the Willamette River. Hard pressing with Google Maps will also give you the lat/lon in decimal degrees. 

```{r}
portland <- data.frame(y = 45.526104, 
                       x = -122.666917) %>% 
            st_as_sf(coords = c("x", "y")) # chose the same coordinate system as above
```

Create a buffer around the chosen point to clip the road data to. The data should really be in a projected coordinate system but this approach still works. The `dist` value here is arc degrees which is a difficult unit to use, but its essentially a distance value. Projecting the data to a UTM coordinate system is probably a better approach.

Check the buffer distance. This was an iterative process to get the `dist` value correct made easier with `{Leaflet}`.



```{r warning=FALSE, message=FALSE}
buf <- st_buffer(portland, dist = .03)

# l <- 
  leaflet(buf) %>%
  addPolygons() %>% 
  addTiles()

# widgetframe::frameWidget(l, width = "100%")
```
*There may be some warnings about the coordinate system being in lat/long and the distance being in degrees*

Crop the roads layer once the buffer area looks good to you.

```{r}
buf <- st_set_crs(buf, 'WGS84')

roads <- st_intersection(roads, buf) 
```

There are lots of suffixes that are not well represented in the data. To keep the legend a reasonable size I'm selecting only the top seven suffixes.

```{r}
top_seven_roads <-
  roads %>%
  as.data.frame() %>% # drop geometry column
  count(FTYPE) %>%
  arrange(desc(n)) %>% # arrange from most to least
  slice(1:7) %>% # take the top seven
  pull(FTYPE)

top_seven_roads
```

Again to keep the legend a reasonable size we set all other suffixes to 'Other'. I'm also going to reclass `RAMP` as other. This is an aesthetic choice since on and off ramps aren't really roads and I want to keep the legend more manageable.

```{r}
roads <- roads %>%
    mutate(FTYPE = case_when(FTYPE == "RAMP" ~ "Other",
                             FTYPE %in% top_seven_roads ~ FTYPE,
                                  TRUE ~ "Other"))
roads$FTYPE %>%
  unique()
```

## Prep data for plotting

Create a vector of colors that is associated with each road type and reclassify the characters to factors. This allows the legend to be in the desired order.

```{r}
# convert to factor for proper ordering in the final plot
# There is a tidy way of doing this, this method is more natural for me
roads$FTYPE <- factor(roads$FTYPE,
                           levels = c("AVE",
                                      "ST",
                                      "FWY",
                                      "BLVD",
                                      "RD",
                                      "PKWY",
                                      "Other"))
```


## Plotting

Plot out the finished product. I put the most time into this part and there are probably other ways to achieve the same results with different ggplot functions. I've added comments to help identify what does what. You may have to change the `plot.title` depending on the output size of the png.

I used the `{cartography}` package for colors here and I would say color choice makes a large difference in the final product. There is plenty of room to play around with the output.

**Note: Since first creating this ggplot has been throwing an odd error when I went to recreate in Jan 2022. ggplot can be mostly replaced with a simple `plot()` call**
```{r}
# ggplot() +
#   geom_sf(data = roads,
#           aes(colour = FTYPE), # color by column value
#           show.legend = "point") + # get legend as circles
#   theme_void() + # remove default gray background
#   theme(legend.position = "bottom",
#         text = element_text(size = 20)) + # set all text large. Play with this number based on desired img size
#   guides(color = guide_legend(nrow = 2, # have legend all on one line
#                               override.aes = (list(size = 5)))) + # change size of point symbols in legend
#   scale_color_manual(name = "Street Suffix", # name of legend
#                      values = carto.pal("pastel.pal")[seq(1,14,2)]) + # our selected color palette
#   ggtitle("Portland Roads") +
#   theme(plot.title = element_text(size = 32))

plot(roads, main = "Portland Roads", key.pos = 1, key.width = .15)
```

[Final Map](/post/Stylized-Street-Maps/portlandstreets.png)

## Save output

```{r eval=F}
# ggsave("portlandstreets.png",
#        width = 12,
#        height = 18,
#        units = "in", # specify units of above height and width
#        dpi = 400) # dots per inch
```

This workflow creates a nice stylized map of street suffixes for Portland that can be extended to any area where you have road data.


